Global phase diagram and possible quantum spin liquid in the triangular $J$-$K$-$\Gamma$ model

### 2020-11-11 paper/paper.tex
Paper 2020-08-18 14:20

\begin{abstract}
The celebrated Kitaev honeycomb model provides an analytically tractable example with an exact quantum spin liquid ground state. While in real materials, other types of interactions besides the Kitaev coupling are present and these models can also be generalized to a triangular lattice. Here, we study the classical and quantum Heisenberg-Kitaev-$\Gamma$ model on a triangular lattice using the classical Monte Carlo and exact diagonalization. We find that the quantum phase diagram is quite similar to its classical counterpart in the large region, suggesting that the Kitaev physics would have emerged largely in the classical limit. Near the antiferromagetic (AF) Kitaev point, we show that the stripe order is selected out of the classic nematic order via the order-by-disorder mechanism. Around the AF $\Gamma$ point, we find a possible quantum spin liquid in a noticeable region with a continuum in spin excitations.
\end{abstract}

@@

In this paper, we study the global phase diagram and the phase transitions of the triangular lattice $J$-$K$-$\Gamma$ model in detail using a combination of ED and classical analyses. We find that the ED results are basically consistent with the classical analyses such as the classical Monte Carlo simulations. In the HK limit ($\Gamma=0$), there are five classical phases: one FM, two stripe, one $120^{\circ}$ N\'{e}el and its dual phases. For the pure AFM Kitaev model, although the system has highly degenerate classical GSs composed of the stripe and nematic states, the order-by-disorder mechanism caused by quantum fluctuations makes the system select the stripe state to be the GS. When the $\Gamma$ term is included, we find five new phases: two FM, one stripe, one modulated stripe and one possible QSL phases. With the aid of classical analysis, we determine the spin directions for each magnetically ordered phase and explain that why there are phase transitions between the phases with the same type of classical orders. On the other hand, we find that the order-by-disorder mechanism makes the GSs of the pure $\Gamma$ models have FM orders, although the classical analyses suggest that the GSs are highly degenerate. For the possible QSL phase, based on the investigation of the spin excitation spectrum, we find that the GS is probably a gapped $Z_{2}$ QSL.

%The rest of the paper is organized as follows: In Sec.~\ref{sec:Model} we introduce the $J$-$K$-$\Gamma$ model as well as the four-sublattice-transformation that can be applied to the HK limit. The ED and classical Monte Carlo methods are introduced in Sec.~\ref{sec:Methods}. Section~\ref{sec:Results} gives the main results of this paper.

@@

%\section{\label{sec:Methods}Methods}

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.95\textwidth]{fig/QuantumGlobalPhaseDiagram.pdf}
    \caption{\label{fig:QuantumPhaseDiagram}(Color online) Global phase diagram of the triangular lattice $J$-$K$-$\Gamma$ model. The angle $\alpha$ and $\beta$ denote the radial and azimuthal angles, respectively. There are in total ten phases including three FM phases denoted as FM-A, FM-B and FM-C, three stripe phases designated as Stripe-A, Stripe-B and Stripe-C, a modulated stripe phase, a 120$^\circ$ N\'{e}el phase, a dual N\'{e}el phase and a possible QSL phase. Phase boundaries are determined by the singularities of $-\partial^2E_0/\partial\alpha^2$ and $-\partial^2E_0/\partial\beta^2$ from ED calculations except for that depicted by the dashed line between the modulated stripe and stripe-B phases given by the classical analyses.}
\end{figure*}

@@

%The stripe orders are indeed possible GS spin configurations, but there are also many nematic states with the same energy as it at the classical level since the AFM chains are decoupled (see Appendix~\ref{apx:DegeneratedStates} for details). When we consider the quantum fluctuations generated by the Kitaev interactions not along the AFM chains, the $4^{th}$ order corrections will give an effective coupling between the next-nearest-neighbor spins~\cite{PhysRevB.92.184416}, which can stabilize the stripe order. Thus, the degeneracy of the spin configurations is lifted by the order-by-disorder mechanism, and we can identify that the phase containing the AFM Kitaev point of the HK model is a stripe phase, which we label the stripe-C in Fig.~\ref{fig:QuantumPhaseDiagram}. Moreover, every stripe order in this phase can be transformed to another degenerate stripe pattern in the same phase by a FST, so these stripe orders are consistent with the FST.

%However, we can not identify the magnetic properties of the phase containing the AFM Kitaev point ($J=\Gamma=0$, $K=1$), since the points in this phase are still mapped to the points in the same phase under the FST.

% When $\beta=\pi - \arctan\frac{1}{2}$ in Eq.~\eqref{eq:FST}, the FST maps the Hamiltonian~\eqref{eq:Hamiltonian} to FM Heisenberg model of the rotated spin operators $\bm{S}^{\prime}$, whose GS is a FM state. Therefore, transformating back to the original spin basis, a collinear stripe order can be obtained. Similarly, a noncollinear spiral order (Dual N\'{e}el) is obtained from the 120$^\circ$ N\'{e}el order for $\beta=-\arctan\frac{1}{2}$. However, for the phase containing the AFM Kitaev point ($J=\Gamma=0$, $K=1$), since it is mapped to itself under FST, we can not identify its magnetic properties with the FST.

%The properties of each quantum phases, especially the spin configuration of the ordered phases, can be inferred from the SSF defined in Eq.~\eqref{eq:SSF}, so we calculate the SSF for the case of pure AFM Kitaev limit ($J=\Gamma=0$, $K=1$) and present the result in Fig.~\ref{fig:StructureFactors}(a). We find that the SSF shows obvious peaks at the $\tilde{M}$ points, which is a typical characteristic of the stripe order as shown in Fig.~\ref{fig:StructureFactors}(i).

@@

%To reveal the natures of these new phases, we further study the SSFs in these new phases. Since the stripe phase for $K<0$ in the HK limit is located inside a phase of the global phase diagram, we infer this phase to have a stripe order and denote it as the stripe-A phase in Fig.~\ref{fig:QuantumPhaseDiagram}. To confirm this, we present the SSFs for two typical points, i.e. $\alpha=0.3\pi$, $\beta=0.75\pi$ and $\alpha=0.75\pi$, $\beta=0.25\pi$, in the $\Gamma>0$ and $\Gamma<0$ regions in Fig.~\ref{fig:StructureFactors}(b) and (c), respectively. Both of the SSFs show obvious peaks at the $\tilde{M}$ points, which is a signature feature of the stripe phase. In Fig.~\ref{fig:StructureFactors}(d), we illustrate the SSF for $\alpha=0.3\pi$, $\beta=0$, which also shows distinct peaks at the $\tilde{M}$ points, so the phase containing this point is also a stripe phase and we denote it the stripe-B in the phase diagram.

@@

For the other two phases with large areas and mainly located in the $J<0$ region of the phase diagram, we show the SSFs of two representative points with $\alpha=0.3\pi, \beta=1.3\pi$ and $\alpha=0.7\pi, \beta=1.65\pi$ in Fig.~\ref{fig:StructureFactors}(f) and (g), respectively. Since the SSFs are peaked at $\tilde{\Gamma}$ point of the first BZ, so both phases have ferromagnetic orders and they are denoted with FM-B and FM-C in Fig.~\ref{fig:QuantumPhaseDiagram}. We note that there are two special points with $\Gamma=\pm1$ ($J=K=0$) in the two FM phases, which are located at the phase boundary in the classical phase diagram~\cite{PhysRevB.92.165108}. Thus, it is worthy to discuss the properties of the pure $\Gamma$ models in details, and we perform the numerical simulations by using the ``basin-hopping" global optimization and classical Monte Carlo methods. For $\Gamma=+1$, apart from the FM GS with the spin orientation lying in the lattice plane, we also found disordered states having the same energy as the FM state. For $\Gamma=-1$, in addition to the FM order perpendicular to the lattice plane, other long-range ordered states which are energetically degenerate with the FM state also appear (see Appendix~\ref{apx:DegeneratedStates} for details). However, for the quantum model, quantum flucations select the FM ordered states as the GS out of the degenerate manifolds of classical states according to our ED calculations. This selection of states among the degenerate classical ground states is the so the called order-by-disorder mechanism, which has previously been applied in a number of insulating magnets, especially in cases where frustration leads to a degenerate manifold of classical ground state configurations that is broken by quantum fluctuations~\cite{JPSJ.54.4494,PhysRevLett.62.2056,PhysRevLett.88.067203,PhysRevB.81.214419,PhysRevLett.109.077204}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/StructureFactors.pdf}
    \caption{\label{fig:StructureFactors}(Color online) (a)-(h) Representative SSFs from the ED calculations for different interaction parameters. The green dashed lines marks the first BZ corresponding to the triangular lattice. (i) Characteristic wave vectors for the FM and stipe phases. See Appendix~\ref{apx:SSF} for the SSFs of the 120$^\circ$ N\'{e}el phase, dual N\'{e}el phase and pure $\Gamma$ models.}
\end{figure}

@@

In order to confirm the consistency between the above classical analyses and the ED results, we use the spin coherent state to extract the moment direction of these FM phases from our ED cluster GS~\cite{PhysRevB.94.064435}. Since the cluster spin coherent state defined in Eq.~(\ref{eq:Spin-1/2CoherentState}) is captured only by a single pair of $(\theta, \phi)$ for the collinear states, it is easy to determine the direction of the FM orders by inspecting the probability map $P(\theta, \phi) = | \langle \Psi (\theta, \phi) | GS \rangle |^2$. As shown in Fig.~\ref{fig:Proabilities}(a), for the FM-B phase, the peaks of the probability form a ring, which indicates that the moment being constrained to a plane with all directions degenerate. To compare with the classical results, we plot the classical magnetic moment directions with the green dashed line in Fig.~\ref{fig:Proabilities}(a), and we find that the the probability is concentrated near the classical magnetic moment directions. For the FM-C phase, the probability map is clearly peaked at specific directions, which are also consistent with the classical results marked by the green solid circle in Fig. \ref{fig:Proabilities}(b). Moreover, the large overlaps between the exact cluster GSs and FM cluster spin coherent states again provide solid evidences that the ED results are consistent with the classical analyses.

%{\color{red}In the HK limit, for the pure FM Heisenberg model ($J=-1$), the FM state is spin rotational invariant and the ordered moment can point any direction. The Kitaev interaction breaks the spin rotational symmetry and pins the orderings to the axis direction (i.e., $x$, $y$ and $z$ directions) when the quantum fluctuations are included, although the classical energy in Eq.~(\ref{eq:EcFM}) is independent of the moment direction (see Appendix~\ref{apx:LSWOfFMA} for details). When the $\Gamma$ interactions are introduced, positive $\Gamma$ couplings drive the ordered moment to the lattice plane and negative $\Gamma$ couplings pin the ordered moment to be perpendicular to the lattice plane. Therefore, the line with $\Gamma=0$ is the phase boundary of the FM-A and FM-C phases.}

%The resulting probability maps for two representative interaction parameters of FM-B and FM-C phases are shown in Fig.~\ref{fig:Proabilities}(a) and (b), form which we can see that both of the probabilities are concentrated near the classical magnetic moment directions. This large overlap between the exact cluster GS and FM cluster spin coherent state again provides solid evidence that the corresponding regions are FM phases. For the FM-B phase, the probability map reveals the moment being constrained to the lattice plane (marked by the green dashed line in Fig.~\ref{fig:Proabilities}(a)) with all directions degenerate.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/Probabilities.pdf}
    \caption{\label{fig:Proabilities}(Color online) Map of the probabilities of the cluster spin coherent state given by Eq.~\eqref{eq:ClusterCoherentState} in exact cluster GS for (a) FM-A, (b) FM-C, (c) stripe-A and (d) stripe-B phases. The radial and polar coordinate gives the angles $\theta$ and $\phi$, which are spherical angles with respect to the reference frame shown in Fig.~\ref{fig:ModelDefinition}(b). The green dashed lines and solid circles mark the ordered moment directions for the classical states. Note that the GSs are six-fold degenerate for the stripe-A and stripe-B phase, and two-fold degenerate for the FM-C phase, so the total probabilities for all of these phases are approximate to $1$. }
\end{figure}

@@

% The following are the Appendixes of this article
%\newpage

@@

%\section{\label{apx:LSWOfFMA}Effects of Quantum fluctuations on FM phases}

%According to Eq.~(\ref{eq:EcFM}), the spin directions in the FM state are independent of the Kitaev interactions, but here we demonstrate that the Kitaev interaction can make the spins point to some specific directions when the quantum fluctuations are included by a linear spin-wave analysis. Within the linear spin-wave theory, the GS energy per lattice site is given by $E_{SW}(\theta, \phi) = E^{c} + 1/(2N) \sum_{\bm{k}} \omega_{\bm{k}}(\theta, \phi)$, where $\omega_{\bm{k}}(\theta, \phi)$ is the spin-wave dispersion. The angles $\theta \in [0, \pi]$ and $\phi \in [0, 2\pi]$ are direction angles of the spins in spherical coordinate. The specific directions are selected if the minimum value of $E_{SW}(\theta, \phi)$ is arrived with some parameters $(\theta, \phi)$. For the FM states of the $J$-$K$-$\Gamma$ model, the dispersion of magnon $\omega_{\bm{k}}(\theta, \phi)$ is given by
%\begin{equation}
%    \omega_{\bm{k}}(\theta, \phi)= \frac{1}{2} \sqrt{A_{\bm{k}}^{2}(\theta, \phi) - |B_{\bm{k}}(\theta, \phi)|^{2}}
%    \label{eq:LSWDispersion}
%\end{equation}
%where
%\begin{equation}
%A_{\bm{k}}(\theta, \phi)=A_{0} + \eta_{1} A_{1} + \eta_{2} A_{2} + \eta_{3} A_{3}, \nonumber
%\end{equation}
%\begin{equation}
%A_{0}=-6J - 2K - 4 \Gamma (v_{x} v_{z} + v_{y} v_{z} + v_{x} v_{y}), \nonumber
%\end{equation}
%\begin{equation}
%A_{1}=2J + K (1-v_{x}^{2}) - 2 \Gamma v_{y} v_{z}, \nonumber
%\end{equation}
%\begin{equation}
%A_{2}=2J + K (1-v_{z}^{2}) - 2 \Gamma v_{x} v_{y}, \nonumber
%\end{equation}
%\begin{equation}
%A_{3}=2J + K (1-v_{y}^{2}) - 2 \Gamma v_{x} v_{z}, \nonumber
%\end{equation}
%and
%\begin{equation}
%B_{\bm{k}}(\theta, \phi)=\eta_{1} B_{1} + \eta_{2} B_{2} + \eta_{3} B_{3}, \nonumber
%\end{equation}
%\begin{align}
%B_{1}=K (1 - v_{x}^{2}) + 2\Gamma v_{y} v_{z}- 2\text{i} (K \frac{v_{x} v_{y} v_{z}}{1-v_{z}^{2}} + \Gamma v_{x}), \nonumber
%\end{align}
%\begin{align}
%B_{2}=K (1 - v_{z}^{2}) + 2 \Gamma \frac{1 + v_{z}^{2}}{1 - v_{z}^{2}} v_{x}v_{y}+ 2\text{i} \Gamma \frac{v_{z}}{1 - v_{z}^{2}}(v_{x}^2 - v_{y}^2), \nonumber
%\end{align}
%\begin{align}
%B_{3}=- K (1 - v_{y}^{2})- 2 \Gamma v_{x}v_{z}+ 2\text{i} (K \frac{v_{x} v_{y} v_{z}}{1-v_{z}^{2}} + \Gamma v_{y}). \nonumber
%\end{align}
%Here $\eta_{1,2,3} = \cos \bm{k} \cdot \bm{a}_{1,2,3}$ and $\bm{a}_{1,2,3}$ are the three nearest neighbor bonds, $v_x$, $v_y$, $v_z$ are the three components of the classical spin with $v_{x} = \sin \theta \cos \phi$, $v_{y} = \sin \theta \sin \phi$ and $v_{z} = \cos \theta$. In the HK limit ($\Gamma = 0$), the first derivatives of $\omega_{\bm{k}}(\theta, \phi)$ versus $\theta$ and $\phi$ are equal to zero when $v_{x} = \pm1$, $v_{y} = \pm1$ or $v_{z} = \pm1$. Accordingly, $E_{SW}(\theta, \phi)$ takes extreme value at these special moment directions. Combined with numerical calculations, the minimum value of $E_{SW}(\theta, \phi)$ is obtained when the direction of the magnetic moment is along $x$, $y$, $z$ axis direction. Thus, quantum fluctuation lift the degeneracy of the classical GS and select FM states with specific moment directions (i.e., $x$, $y$ and $z$ directions) as the quantum GSs.

>>>>

\begin{abstract}
The celebrated Kitaev honeycomb model provides an analytically tractable example with an exact quantum spin liquid ground state. While in real materials, other types of interactions besides the Kitaev coupling are present and these models can also be generalized to a triangular lattice. Here, we study the classical and quantum Heisenberg-Kitaev-$\Gamma$ model on a triangular lattice using the classical Monte Carlo and exact diagonalization. We find that the quantum phase diagram is quite similar to its classical counterpart in the large region, suggesting that the Kitaev physics would have emerged largely in the classical limit. Near the antiferromagetic Kitaev point, we show that the stripe order is selected out of the classic nematic order via the order-by-disorder mechanism. Also, the order-by-disorder mechanism selects the ferromagnetic ordered states as the ground state out of the degenerate classical states for the pure $\Gamma$ models. Around the positive $\Gamma$ point, due to the interplay of the Heisenberg, Kiatev and $\Gamma$ interactions, we find a possible quantum spin liquid in a noticeable region with a continuum in spin excitations.
\end{abstract}

@@

In this paper, we comprehensively study the global phase diagram and the phase transitions of the triangular lattice $J$-$K$-$\Gamma$ model using a combination of ED and classical analyses. We find that the ED results are basically consistent with the classical analyses such as the classical Monte Carlo simulations. In the HK limit ($\Gamma=0$), there are five classical phases: one FM, two stripe, one $120^{\circ}$ N\'{e}el and its dual phases. For the pure AFM Kitaev model, although the system has highly degenerate classical GSs composed of the stripe and nematic states, the order-by-disorder mechanism caused by quantum fluctuations makes the system select the stripe state to be the GS. When the $\Gamma$ term is included, we find five new phases: two FM, one stripe, one modulated stripe and one possible QSL phases. With the aid of classical analysis, we determine the spin configuration for each magnetically ordered phase and explain that why there are phase transitions between the phases with the same type of classical orders. On the other hand, we find that the order-by-disorder mechanism makes the GSs of the pure $\Gamma$ models have FM orders, although the classical analyses suggest that the GSs are highly degenerate. For the possible QSL phase, based on the investigation of the spin excitation spectrum, we find that the GS is probably a gapped $Z_{2}$ QSL.

@@

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.95\textwidth]{fig/QuantumGlobalPhaseDiagram.pdf}
    \caption{\label{fig:QuantumPhaseDiagram}(Color online) Global phase diagram of the triangular lattice $J$-$K$-$\Gamma$ model. The angle $\alpha$ and $\beta$ denote the radial and azimuthal angles, respectively. There are in total ten phases including three FM phases denoted as FM-A, FM-B and FM-C, three stripe phases denoted as Stripe-A, Stripe-B and Stripe-C, a modulated stripe phase, a 120$^\circ$ N\'{e}el phase, a dual N\'{e}el phase and a possible QSL phase. Phase boundaries are determined by the singularities of $-\partial^2E_0/\partial\alpha^2$ and $-\partial^2E_0/\partial\beta^2$ from ED calculations except for that depicted by the dashed line between the modulated stripe and stripe-C phases given by the classical analyses.}
\end{figure*}

@@

@@

@@

For the other two phases with large areas and mainly located in the $J<0$ region of the phase diagram, we show the SSFs of two representative points with $\alpha=0.3\pi, \beta=1.3\pi$ and $\alpha=0.7\pi, \beta=1.65\pi$ in Fig.~\ref{fig:StructureFactors}(f) and (g), respectively. Since the SSFs are peaked at $\tilde{\Gamma}$ point of the first BZ, so both phases have ferromagnetic orders and they are denoted with FM-B and FM-C in Fig.~\ref{fig:QuantumPhaseDiagram}. We note that there are two special points with $\Gamma=\pm1$ ($J=K=0$) in the two FM phases, which are located at the phase boundary in the classical phase diagram~\cite{PhysRevB.92.165108}. Thus, it is worthy to discuss the properties of the pure $\Gamma$ models in details, and we perform the classical analyses by using the ``basin-hopping" global optimization and classical Monte Carlo methods. For $\Gamma=+1$, apart from the FM GS with the spin orientation lying in the lattice plane, we also found disordered states having the same energy as the FM state (see Appendix~\ref{apx:DegeneratedStates} for details). For $\Gamma=-1$, in addition to the FM order perpendicular to the lattice plane, there are also several noncollinear magnetic orders energetically degenerate with the FM state (see Appendix~\ref{apx:DegeneratedStates} for details). However, according to our ED calculations, for the quantum model, quantum flucations select the FM ordered states as the GS out of the degenerate manifolds of classical states. This selection of states among the degenerate classical ground states is the so the called order-by-disorder mechanism, which has previously been applied in a number of insulating magnets, especially in cases where frustration leads to a degenerate manifold of classical ground state configurations that is broken by quantum fluctuations~\cite{JPSJ.54.4494,PhysRevLett.62.2056,PhysRevLett.88.067203,PhysRevB.81.214419,PhysRevLett.109.077204}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/StructureFactors.pdf}
    \caption{\label{fig:StructureFactors}(Color online) SSFs from the ED calculations for representative interaction parameters in different phases: (a) strip-B, (b) and (c) stripe-A, (d) stripe-C, (e) modulated stripe, (f) FM-B, (g) FM-C, (h) QSL. The green dashed lines marks the first BZ of the triangular lattice. (i) Characteristic wave vectors for the FM and stipe phases. See Appendix~\ref{apx:SSF} for the SSFs of the 120$^\circ$ N\'{e}el phase, dual N\'{e}el phase and pure $\Gamma$ models.}
\end{figure}

@@

In order to confirm the consistency between the above classical analyses and the ED results, we use the spin coherent state to extract the moment direction of these FM phases from our ED cluster GS~\cite{PhysRevB.94.064435}. Since the cluster spin coherent state defined in Eq.~(\ref{eq:Spin-1/2CoherentState}) is captured only by a single pair of $(\theta, \phi)$ for the collinear states, it is easy to determine the direction of the FM orders by inspecting the probability map $P(\theta, \phi) = | \langle \Psi (\theta, \phi) | GS \rangle |^2$. As shown in Fig.~\ref{fig:Proabilities}(a), for the FM-B phase, the peaks of the probability form a ring, which indicates that the moment being constrained to a plane with all directions degenerate. To compare with the classical results, we plot the classical magnetic moment directions with the green dashed line in Fig.~\ref{fig:Proabilities}(a), and we find that the the probability is concentrated near the classical magnetic moment directions. For the FM-C phase, the probability map is clearly peaked at specific directions, which are also consistent with the classical results marked by the green solid circle in Fig. \ref{fig:Proabilities}(b). Moreover, the large overlaps between the exact cluster GSs and FM cluster spin coherent states again provide solid evidences that the ED results are consistent with the classical analyses.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/Probabilities.pdf}
    \caption{\label{fig:Proabilities}(Color online) Maps of the probabilities of the cluster spin coherent states given by Eq.~\eqref{eq:Spin-1/2CoherentState} in exact cluster GS for (a) FM-B, (b) FM-C, (c) stripe-A and (d) stripe-C phases. The radial and polar coordinate gives the angles $\theta$ and $\phi$, which are spherical angles with respect to the reference frame shown in Fig.~\ref{fig:ModelDefinition}(b). The green dashed lines and solid circles mark the ordered moment directions for the classical states. Note that the GSs are two-fold degenerate for the FM-C phase, and six-fold degenerate for the stripe-A and stripe-C phases, so the total probabilities for all of these phases are approximate to $1$. }
\end{figure}

@@

@@
